---
title: "Hands-On_Ex04 - Spatio-Temporal Point Patterns Analysis"
title-block-banner: true
author: "Han Ming Yan"
date: 2024-09-09
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    number-depth: 3
    code-copy: true
    embed-resources: true
    lightbox: true
    lang: en
---

## **Overview**

A spatio-temporal point process (also called space-time or spatial-temporal point process) is a random collection of points, where each point represents the time and location of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years. This chapter shows how various R packages can be combined to run a set of spatio-temporal point pattern analyses in a guided and intuitive way. A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

## **Learning Outcome**

### **The research questions**

The specific questions we would like to answer are:

-   are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?

-   if the answer is NO, where and when the observed forest fire locations tend to cluster?

## **The data**

For the purpose of this exercise, two data sets will be used, they are:

-   *forestfires*, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from [Fire Information for Resource Management System](https://firms.modaps.eosdis.nasa.gov/download/). For the purpose of this exercise, only forest fires within Kepulauan Bangka Belitung will be used.

-   *Kepulauan_Bangka_Belitung*, an ESRI shapefile showing the sub-district (i.e. [kelurahan](https://en.wikipedia.org/wiki/Villages_of_Indonesia)) boundary of Kepulauan Bangka Belitung. The data set was downloaded from [Indonesia Geospatial](https://www.indonesia-geospasial.com/2023/05/download-shapefile-batas-administrasi.html) portal. The original data covers the whole Indonesia. For the purpose of this exercise, only sub-districts within Kepulauan Bangka Belitung are extracted.

## **Installing and Loading the R packages**

For the purpose of this study, six R packages will be used. They are:

-   [**sf**](https://r-spatial.github.io/sf/) provides functions for importing processing and wrangling geospatial data,,

-   [**raster**](https://rspatial.org/raster/) for handling raster data in R,

-   [**spatstat**](https://cran.r-project.org/web/packages/spatstat/index.html) for performing Spatial Point Patterns Analysis such as kcross, Lcross, etc.,

-   [**sparr**](https://tilmandavies.github.io/sparr/index.html) provides functions to estimate fixed and adaptive kernel-smoothed spatial relative risk surfaces via the density-ratio method and perform subsequent inference. Fixed-bandwidth spatiotemporal density and relative risk estimation is also supported

-   [**tmap**](https://r-tmap.github.io/tmap/) provides functions to produce cartographic quality thematic maps, and

-   [**tidyverse**](https://www.tidyverse.org/), a family of R packages that provide functions to perform common data science tasks including and not limited to data import, data transformation, data wrangling and data visualisation.

## Load the R packages using Pacman::p_load

```{r}
pacman::p_load(sf, raster, spatstat, sparr, tmap, tidyverse)
```

## **Importing and Preparing Study Area**

### **Importing study area**

Code chunk below is used import study area (i.e. Kepulauan Bangka Belitung) into R environment.

```{r}
kbb <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

In sf format:

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

### **Converting OWIN**

Next, `as.owin()` is used to convert *kbb* into an **owin** object.

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

Use `class()` is used to confirm if the output is indeed an owin object.

```{r}
class(kbb_owin)
```

## **Importing and Preparing Forest Fire data**

Next, we will import the forest fire data set (i.e. *forestfires.csv*) into R environment.

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```

Because **ppp** object only accept numerical or character as mark. The code chunk below is used to convert data type of *acq_date* to numeric.

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## **Visualising the Fire Points**

### **Overall plot**

We will now prepare a point symbol map showing the distribution of fire points using this code chunk below.

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

### **Visuaising geographic distribution of forest fires by month**

Next, we will prepare a point symbol map showing the monthly geography distribution of fire points in 2023 using this code chunk below.

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## **Computing STKDE by Month**

In this section, you will learn how to compute STKDE by using [`spattemp.density()`](https://tilmandavies.github.io/sparr/reference/spattemp.density.html) of **sparr** package. Before using the function, it is highly recommended you read the function’s reference guide in detail in order to understand the input data requirements and the output object generated.

### **Extracting forest fires by month**

The code chunk below is used to remove the unwanted fields from *fire_sf* sf data.frame. This is because `as.ppp()` only need the mark field and geometry field from the input sf data.frame.

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

### **Creating ppp**

The code chunk below is used to derive a **ppp** object called *fire_month* from *fire_month* sf data.frame.

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

Check if the output is in the correct object class

```{r}
summary(fire_month_ppp)
```

Check if there are duplicated point events by using the code chunk below.

```{r}
any(duplicated(fire_month_ppp))
```

### **Including Owin object**

The code chunk below is used to combine *origin_am_ppp* and *am_owin* objects into one.

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

As a good practice, `plot()` is used to plot *ff_owin* so that we can examine the correctness of the output object.

```{r}
plot(fire_month_owin)
```

### **Computing Spatio-temporal KDE**

Next, `spattemp.density()` of sparr package is used to compute the STKDE.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### **Plotting the spatio-temporal KDE object**

In the code chunk below, `plot()` of R base is used to the KDE for between July 2023 - December 2023.

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

## **Computing STKDE by Day of Year**

In this section, you will learn how to computer the STKDE of forest fires by day of year.

### **Creating ppp object**

In the code chunk below, DayofYear field is included in the output ppp object.

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### **Including Owin object**

Next, code chunk below is used to combine the ppp object and the owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

plot the kde_yday which is each day of the year

```{r}
plot(kde_yday)
```

## **Computing STKDE by Day of Year: Improved method**

One of the nice function provides in sparr package is `BOOT.spattemp()`. It support bandwidth selection for standalone spatiotemporal density/intensity based on bootstrap estimation of the MISE, providing an isotropic scalar spatial bandwidth and a scalar temporal bandwidth.

Code chunk below uses `BOOT.spattemp()` to determine both the spatial bandwidth and the scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

### **Computing spatio-temporal KDE**

Now, the STKDE will be derived by using h and lambda values derive in previous step.

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin,
  h = 9000,
  lambda = 19)
summary(kde_yday)
```

### **Plotting the output spatio-temporal KDE**

Last, `plot()` of **sparr** package is used to plot the output as shown below.

```{r}
plot(kde_yday)
```
