---
title: "In-Class Exercise 2: Thematic Mapping and GeoVisualisation with R"
title-block-banner: true
author: "Han Ming Yan"
date: 2024-08-26
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 3
execute: 
  eval: true
  echo: true
  warning: true
  freeze: true
---

## *Working with Master Plan Planning Sub-Zone Data*

::: panel-tabset
### The Task

1.  Visit data.gov.sg and download Master Plan 2019 Subzone Boundary (No Sea) from the portal. You are required to download both kml files.

2.  Move MPSZ-2019 shapefile provided for In-Class Exercise 1 folder on elearn to data sub-folder of In-Class_Ex02.

3.  Write a code chunk to import Master Plan 2019 Subzone Boundary (No Sea) kml and MPSZ-2019 into sf simple feature data.frame.

### To import shapefile

The code chunk below will be used to install and load these packages in RStudio.

::: {style="font-size: 1.50em"}
```{r}
pacman::p_load(sf, tidyverse)
```
:::

```{r}
#| echo: false
pacman::p_load(sf, tidyverse, ggstatsplot, tmap)
```

The code chunk below uses the *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz14_shp`.

```{r}
mpsz19_shp <- st_read(dsn = "data",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

### To import kml

The code chunk below will be used to install and load these packages in RStudio.

::: {style="font-size: 1.50em"}
```{r}
pacman::p_load(sf, tidyverse)
```
:::

```{r}
#| echo: false
pacman::p_load(sf, tidyverse, ggstatsplot, tmap)
```

The code chunk imports kml file. ::: {style="font-size: 1.50em"}

```{r}
#| echo: false
mpsz19_kml <- st_write(mpsz19_shp, 
                       "data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml",
                       delete_dsn = TRUE)
```
:::

### **Import Pre-School Shape**

::: panel-tabset
### Pre-School Shape

```{r}
preschool <- st_write(mpsz19_shp, 
                       "data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml",
                       delete_dsn = TRUE) #incomplete
```
:::

### **Working with Population Data**

::: panel-tabset
### The Code

```{r}
popdata <- readr::read_csv("data/aspatial/respopagesextod2023.csv")
```
:::

::: panel-tabset
### Data Wrangling

```{r}
popdata2023 <- popdata |>
  dplyr::group_by(PA, SZ, AG) |>
  dplyr::summarise(`POP`=sum(`Pop`)) |>
  dplyr::ungroup() |>
  tidyr::pivot_wider(names_from = AG,
              values_from = POP
              )
colnames(popdata2023)
```
:::

::: panel-tabset
### Data Processing

```{r}
popdata2023 <- popdata2023 %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[14])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:13])+
rowSums(.[15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### Mutate the code

```{r}
popdata2023 <- popdata2023 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)

```

### LEFT_JOIN

```{r}
mpsz_pop2023 <- left_join(popdata2023 ,mpsz19_shp,
                          by = c("SZ"="SUBZONE_N"))

write_rds(mpsz_pop2023, "data/rds/mpszpop2023.rds")
```
:::
