---
title: "Take-Home Exercise 1: Geospatial analytics for Social Good - Myanmar Arm Conflict Case Study"
title-block-banner: true
author: "Han Ming Yan"
date: 2024-09-02
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    number-depth: 3
    code-copy: true
    embed-resources: true
    lightbox: true
    lang: en
abstract: |
  In this study, you are tasked to apply spatial point patterns analysis methods to discover the spatial and spatio-temporal distribution of armed conflict in Myanmar.
keywords: ["Spatial Analysis", "Point Patterns", "First-Order Analysis", "Second-Order Analysis", "Monte Carlo simulation", "Myanmar", "Myanmar Civil War Crisis"]
---

# **Overview**

This exercise focuses on applying geospatial analytics to explore and analyze the impact of the armed conflict in Myanmar. The aim is to use spatial data and analytical techniques to better understand the conflict's dynamics, identify affected regions, and assess the humanitarian implications. This case study offers a real-world application of geospatial tools for social good, demonstrating how data-driven insights can inform decision-making in crisis situations.

## **Data used**

For the purpose of this assignment, armed conflict data of Myanmar between 2021-2024 from [Armed Conflict Location & Event Data (ACLED)](https://acleddata.com/), an independent, impartial, international non-profit organization collecting data on violent conflict and protest in all countries and territories in the world, should be used.

In terms of event types, students should focus on at least four main event types, namely: Battles, Explosions/Remote Violence, Strategic developments, and Violence against civilians.

In terms of study period, students should focus on quarterly armed conflict events from January 2021 until June 2024.

-   ACLED_Myanmar.csv in the data folder

Tools used:

-   [**acledata**](https://acleddata.com/data-export-tool/) **requires Access Key, Given**

::: panel-tabset
## The Task

The specific tasks of this take-home exercise are as follows:

-   Using appropriate function of **sf** and **tidyverse** packages, import and transform the downloaded armed conflict data and administrative boundary data into sf tibble data.frames.

-   Using the geospatial data sets prepared, derive quarterly KDE layers.

-   Using the geospatial data sets prepared, perform 2nd-Order Spatial Point Patterns Analysis.

-   Using the geospatial data sets prepared, derive quarterly spatio-temporal KDE layers.

-   Using the geospatial data sets prepared, perform 2nd-Order Spatio-temporal Point Patterns Analysis.

-   Using appropriate **tmap** functions, display the KDE and Spatio-temporal KDE layers on openstreetmap of Myanmar.

-   Describe the spatial patterns revealed by the KDE and Spatio-temporal KDE maps.

## Grading Criteria

-   **Geospatial Data Wrangling (20 marks):** This is an important aspect of geospatial analytics. You will be assessed on your ability to employ appropriate R functions from various R packages specifically designed for modern data science such as readxl, tidyverse (tidyr, dplyr, ggplot2), sf just to mention a few of them, to perform the entire geospatial data wrangling processes, including. This is not limited to data import, data extraction, data cleaning and data transformation. Besides assessing your ability to use the R functions, this criterion also includes your ability to clean and derive appropriate variables to meet the analysis need.

<!-- -->

-   **Geospatial Analysis (30 marks):** In this exercise, you are expected to use the appropriate spatial point patterns analysis methods and R packages introduced in class to analysis the geospatial data prepared. You will be assessed on your ability to derive analytical products by using appropriate kernel estimation techniques.

-   **Geovisualisation and geocommunication (20 marks):** In this section, you will be assessed on your ability to communicate Exploratory Spatial Data Analysis and Confirmatory Spatial Data Analysis results in layman friendly visual representations. This course is geospatial centric, hence, it is important for you to demonstrate your competency in using appropriate geovisualisation techniques to reveal and communicate the findings of your analysis.

-   **Reproducibility (15 marks):** This is an important learning outcome of this exercise. You will be assessed on your ability to provide a comprehensive documentation of the analysis procedures in the form of code chunks of Quarto. It is important to note that it is not enough by merely providing the code chunk without any explanation on the purpose and R function(s) used.

-   **Bonus (15 marks):** Demonstrate your ability to employ methods beyond what you had learned in class to gain insights from the data.
:::

## **Installing and Loading the R packages**

In this hands-on exercise, five R packages will be used, they are:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.

-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.

-   sparr provides functions to estimate fixed and adaptive kernel

-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.

-   [**maptools**](https://cran.r-project.org/web/packages/maptools/index.html) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**. (Deprecated as of October 2023)

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

Use the code chunk below to install and launch the five R packages.

# Installing maptools

**maptools** is retired as of October 2023 and binary is removed from CRAN. However, we can still download from [**Posit Public Package Manager**](https://packagemanager.posit.co/client/) snapshots by using the code chunk specified

```{r}
#|eval: false # it will no longer run anymore after first run
#|warning: false
install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

```{r}
pacman::p_load(tmap, sf, tidyverse, raster, spatstat, sp, sparr, dplyr)
```

Since it is a .csv file, we will use readr::read_csv() to get the Myanmar data. We will also import the kml data which gives the outline of the country

```{r}
acled_sf <- readr::read_csv("data/myanmar/ACLED_Myanmar.csv") %>%
  st_as_sf(coords = c( "longitude","latitude"),
           crs = 4326) %>%
  st_transform(crs = 32647) %>%
    mutate(event_date = dmy(event_date))
#acled_sf <- acled_sf %>% dplyr::filter(event_type == "Political Violence")
adm0 <- st_read("data/myanmar/geonode-mmr_polbnda_adm0_250k_mimu_1.kml")
adm2 <- st_read("data/myanmar/geonode-mmr_polbnda_adm2_250k_mimu.kml")
adm3 <- st_read("data/myanmar/geonode-mmr_polbnda_adm3_250k_mimu_1.kml")
```

```{r}
head(acled_sf)
```

Next i want to find out about what column names are found in the csv data, so that i can add/remove columns if needed.

```{r}
colnames(acled_sf)
```

since the iso, country and region will definitely be the same for the entire data, we will remove unneccessary data to make reading the data simpler by using the select function from the dplyr package.

```{r}
acled_sf <- acled_sf %>% dplyr::select(-c("iso", "region", "country", "notes"))
```

Double check if the columns are removed

```{r}
colnames(acled_sf)
```

```{r}
# adm0_polygons <- st_collection_extract(adm0, "POLYGON")
adm2_polygons <- st_collection_extract(adm2, "POLYGON")
adm3_polygons <- st_collection_extract(adm3, "POLYGON")
```

```{r}
tmap_mode('plot') # interactive view

acled_sf %>% filter(year == 2023 |
                      event_type == 'Battles')

tm_shape(adm2_polygons) + 
  tm_borders(col = "green") + 
tm_shape(adm3_polygons) +
  tm_borders(col = "black") +
tm_shape(adm0) + 
  tm_dots(col = "blue", size = 0.1) + 
tm_shape(acled_sf) + 
  tm_dots(col = "red", size = 0.1) # Plot the conflict data points in red
```

### Save data into the rds folder

We do this so that we do not need to re-run the previous codes, to keep the derived data in a .rds file

```{r}
# | eval: false
write_rds(acled_sf, "data/rds/acled_sf.rds") # keep in tibble format
```

### To read back the data, (if we do not have it in our environment yet)

```{r}
#| echo: false # this allows user to not see the code output lines, but it is still running
acled_sf <- read_rds("data/rds/acled_sf.rds")
```

## Convert acled_sf into their various quarters (14 in total)

```{r}
# Convert event_date to quarters for quarterly analysis, shift quarter to 4th column for easier reference
myanmar_sf <- acled_sf %>%
  mutate(quarter = paste0(year(event_date), " Q", quarter(event_date))) %>%
  dplyr::select(1:3, quarter, everything())
```

### Extract into various data frames based on the quarters

we do this so that we have lesser data to work with for each data frame, making it faster overall

### Split the quarters based on the event_type in the myanmar_sf data frames

```{r}
# Get all quarters from the data
unique_quarters <- unique(myanmar_sf$quarter)
event_types <- c(
  "Battles" = "b",
  "Strategic developments" = "sd",
  "Violence against civilians" = "vac",
  "Explosions/Remote violence" = "erv"
)

for (q in unique_quarters) {
  
  df_quarter <- myanmar_sf %>% filter(quarter == q)
  
  for (event in names(event_types)) {
    df_event <- df_quarter %>% filter(event_type == event)
    
    # for debugging:
    #print(paste0("Processing quarter: ", q, ", event: ", event, " - Rows: ", nrow(df_event)))
    
    df_name <- paste0("myanmar_sf_", gsub(" ", "_", q), "_", event_types[[event]])
    
    assign(df_name, df_event)
  }
}

ls(pattern = "myanmar_sf_")
```

Since I have split the data frames into their respective year, quarter and categories, i will remove the original quarter only data frames and also remove all the empty data frames (0 rows.)

```{r}
old_quarter_dfs <- ls(pattern = "^myanmar_sf_\\d{4}_Q\\d$")

# Remove the old quarterly data frames
rm(list = old_quarter_dfs)
all_dfs <- ls(pattern = "^myanmar_sf_")

# Loop through each data frame and remove those with 0 rows
for (df_name in all_dfs) {
  df <- get(df_name)
  if (nrow(df) == 0) {
    rm(list = df_name)
  }
}
myanmar_sfs <- ls(pattern = "myanmar_sf_")
```

## **Converting myanmar_sf frame to sp’s Spatial\* class**

since we have 14 quarters to convert, instead of converting all one by one, i will use a for-each loop to convert them in one code chunk

```{r}
myanmar_sfs <- ls(pattern = "^myanmar_sf_")

# Loop through each of these data frames and convert to Spatial objects
for (df_name in myanmar_sfs) {
  df <- get(df_name)
  sp_object <- as_Spatial(df)
  sp_name <- gsub("sf", "sp", df_name)
  assign(sp_name, sp_object)
}

myanmar_sps <- ls(pattern = "myanmar_sp_")
myanmar_sps
```

## Display one quarter to check

In this case i used myanmar_sp_2023_Q3_sd

```{r}
myanmar_sp_2023_Q3_sd
```

## **Converting the generic sp format into spatstat’s ppp format**

```{r}
for (df_name in myanmar_sfs) {
  df <- get(df_name)
  # Convert to ppp object
  ppp_object <- as.ppp(df)
  
  # Create a name for the new ppp object
  ppp_name <- gsub("sf", "ppp", df_name)
  
  # Assign the ppp object to a variable with the new name
  assign(ppp_name, ppp_object)
}
myanmar_ppps <- ls(pattern = "myanmar_ppp_")
myanmar_ppps
```

```{r}
myanmar_ppp_2023_Q3_b
plot(myanmar_ppp_2023_Q3_b)
```

```{r}
summary(myanmar_ppp_2023_Q3_b)
```

```{r}
any(duplicated(myanmar_ppp_2023_Q3_b))
```

```{r}
sum(multiplicity(myanmar_ppp_2023_Q3_b) > 1)
```

### **Creating *owin* objects**

```{r}
for (df_name in myanmar_sfs) {
  ppp_name <- gsub("sf", "ppp", df_name)
  ppp_object <- get(ppp_name)
  owin_object <- as.owin(ppp_object)
  owin_name <- gsub("sf", "owin", df_name)
  assign(owin_name, owin_object)
}
ls(pattern = "myanmar_owin_")
```

```{r}
plot(myanmar_owin_2024_Q1_vac)
```

```{r}
summary(myanmar_owin_2024_Q1_vac)
```

# Second-order Spatial Point Patterns Analysis

### **G-Function Analysis**

The G-function measures the distribution of the distances from an arbitrary event to its nearest event.

```{r}
# Loop through each quarterly ppp object to compute G-function
for (ppp_name in myanmar_ppps) {
  # Get the ppp object
  ppp_object <- get(ppp_name)
  
  # Compute G-function estimation
  G <- Gest(ppp_object, correction = "border")
  
  # Plot the G-function
  plot(G, xlim = c(0, 500), main = paste("G-Function for", ppp_name))
  
  # Perform Monte Carlo test with G-function
  G_csr <- envelope(ppp_object, Gest, nsim = 999)
  
  # Plot the Monte Carlo test results
  plot(G_csr, main = paste("CSR Test - G-Function for", ppp_name))
}

```

### **F-Function Analysis**

The F-function estimates the empty space function F(r) from a point pattern in an arbitrary window shape.

```{r}
# Loop through each quarterly ppp object to compute F-function
for (ppp_name in myanmar_ppps) {
  # Get the ppp object
  ppp_object <- get(ppp_name)
  
  # Compute F-function estimation
  F <- Fest(ppp_object)
  
  # Plot the F-function
  plot(F, main = paste("F-Function for", ppp_name))
  F_csr <- envelope(ppp_object, Fest, nsim = 999)
  

  plot(F_csr, main = paste("CSR Test - F-Function for", ppp_name))
}
```

### **K-Function Analysis**

The K-function measures the number of events found up to a given distance of any particular event.

```{r}
# Loop through each quarterly ppp object to compute K-function
for (ppp_name in myanmar_ppps) {
  ppp_object <- get(ppp_name)
  K <- Kest(ppp_object, correction = "Ripley")

  plot(K, . -r ~ r, ylab = "K(d)-r", xlab = "d(m)", main = paste("K-Function for", ppp_name))
  K_csr <- envelope(ppp_object, Kest, nsim = 99, rank = 1, glocal = TRUE)
  plot(K_csr, . - r ~ r, xlab = "d", ylab = "K(d)-r", main = paste("CSR Test - K-Function for", ppp_name))
}
```

### **L-Function Analysis**

The L-function is a linearized version of the K-function, making interpretation easier.

```{r}
# Loop through each quarterly ppp object to compute L-function
for (ppp_name in myanmar_ppps) {
  # Get the ppp object
  ppp_object <- get(ppp_name)
  
  G <- Gest(ppp_object, correction = "border")
  
  plot(G, xlim = c(0, 500), main = paste("G-Function for", ppp_name))
  
  G_csr <- envelope(ppp_object, Gest, nsim = 999)
  
  plot(G_csr, main = paste("CSR Test - G-Function for", ppp_name))
}

```

### **Different Automatic Bandwidth Methods**

Using myanmar_ppp_2021_Q1_Strategic_developments to show case the sigma values for the automatic Bandwidth methods.

```{r}
bw.CvL(myanmar_ppp_2021_Q1_sd)
```

```{r}
bw.scott(myanmar_ppp_2021_Q1_sd)
```

```{r}
bw.ppl(myanmar_ppp_2021_Q1_sd)
```

```{r}
bw.diggle(myanmar_ppp_2021_Q1_sd)
```

| Function  | Methodology                                | Advantages                                      | Limitations                                          |
|------------------|------------------|------------------|-------------------|
| bw.CvL    | Least Squares Cross-Validation             | Data-driven, adaptive                           | Computationally intensive, sensitive to outliers     |
| bw.scott  | Scott’s Rule of Thumb                      | Simple, quick                                   | Assumes Gaussian distribution, non-adaptive          |
| bw.ppl    | Penalized Profile Likelihood               | Balances fit and smoothness, flexible           | Complex, requires penalty parameter tuning           |
| bw.diggle | Diggle’s Method (Edge-corrected, adaptive) | Tailored for spatial data, handles edge effects | More specific, potentially computationally demanding |

-   **For Quick, General Estimates:** `bw.scott()` is suitable when you need a fast, rule-of-thumb bandwidth without deep customization.

-   **For Data-Driven Precision:** `bw.CvL()` offers a more tailored bandwidth selection by minimizing prediction error but at a higher computational cost.

-   **For Balancing Smoothness and Fit:** `bw.ppl()` is ideal when you want to control the trade-off between overfitting and oversmoothing.

-   **For Spatially Inhomogeneous Data with Edge Concerns:** `bw.diggle()` is preferable when dealing with spatial point patterns that exhibit varying intensities and are subject to boundary effects.

```{r}
acled_myanmar_ppl <- density(acled_myanmar_ppp.km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(acled_myanmar.bw, main = "bw.diggle")
plot(acled_myanmar.ppl, main = "bw.ppl")
```

```{r}
par(mfrow=c(2,2))
plot(density(acled_myanmar_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(acled_myanmar_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(acled_myanmar_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(acled_myanmar_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

# Deriving quarterly spatio-temporal KDE layers

### Creating `ppp` Objects with Time Marks

To perform spatio-temporal analysis, we need to create new `ppp` objects that include a temporal component. For this case, we'll use the event_type as the temporal mark.

```{r}

event_type_mapping <- as.numeric(as.factor(myanmar_sf$event_type))
names(event_type_mapping) <- levels(as.factor(myanmar_sf$event_type))

# Convert quarterly data frames to ppp objects with event_type as the mark
for (df_name in ls(pattern = "^myanmar_sf_\\d{4}_Q\\d_")) {
  df <- get(df_name)
  df <- df %>%
    mutate(Event_num = event_type_mapping[as.character(event_type)])
  
  # Convert to ppp object with Event_num as the mark
  ppp_object <- as.ppp(df %>% dplyr::select(Event_num))
  
  assign(gsub("sf", "ppp_temp", df_name), ppp_object)
}

# Check the created ppp objects
ls(pattern = "myanmar_ppp_temp")
```

### Computing Spatio-Temporal KDE

Now, let's use `sparr::spattemp.density()` to compute the spatio-temporal kernel density estimation for each quarterly dataset.

### Function to convert ppp objects to Kilometers (km)

```{r}
# Function to convert ppp objects to kilometers
convert_to_km <- function(ppp_object) {
  print(unitname(ppp_object))
  ppp_km <- rescale.ppp(ppp_object, s = 1000, unitname = c("km", "km"))
  
  return(ppp_km)
}
```

Apply the conversion function to all myanmar_ppp_temp objects

```{r}
ppp_temps <- ls(pattern = "myanmar_ppp_temp")
for (ppp_name in ppp_temps) {
  ppp_object <- get(ppp_name)
  
  # Convert the ppp object to kilometers
  ppp_km <- convert_to_km(ppp_object)
  assign(gsub("temp", "temp_km", ppp_name), ppp_km)
}
ls(pattern = "myanmar_ppp_temp_km")
```

remove all previous ppp_temp data frames

```{r}

all_ppp_objects <- ls(pattern = "^myanmar_ppp_temp_")
non_km_objects <- all_ppp_objects[!grepl("_km_", all_ppp_objects)]
rm(list = non_km_objects)
ls(pattern = "^myanmar_ppp_temp_")

```

List the ppp files into lists based on their event type:

```{r}
b_ppp_list <- list()       # Battles
erv_ppp_list <- list()     # Explosions/Remote Violence
vac_ppp_list <- list()     # Violence Against Civilians
sd_ppp_list <- list()      # Strategic Developments

ppp_object_names <- ls(pattern = "^myanmar_ppp_temp_km_")

for (ppp_name in ppp_object_names) {
  ppp_object <- get(ppp_name)
  
  if (grepl("b", ppp_name)) {
    b_ppp_list[[ppp_name]] <- ppp_object
  } else if (grepl("erv", ppp_name)) {
    erv_ppp_list[[ppp_name]] <- ppp_object
  } else if (grepl("vac", ppp_name)) {
    vac_ppp_list[[ppp_name]] <- ppp_object
  } else if (grepl("sd", ppp_name)) {
    sd_ppp_list[[ppp_name]] <- ppp_object
  }
}

length(b_ppp_list)
length(erv_ppp_list)
length(vac_ppp_list)
length(sd_ppp_list)
```

### Helper Function

You can now use this helper function in your context for each event type (`violence_civilians`, `explosions_remote_violence`, `strat_devs`, `battles`). Below is how you can set it up for each category with the appropriate `ppp_km` objects.

```{r}
combine_ppp_with_owin <- function(ppp_name, ppp_object) {
  # Extract the relevant part of the name to match the owin object
  owin_name <- gsub("ppp_temp_km", "owin", ppp_name)
  
  # Get the corresponding owin object
  owin_object <- get(owin_name)
  
  # Combine the ppp object with the corresponding owin window
  combined_ppp <- ppp_object[owin_object]
  
  return(combined_ppp)
}

plot_combined_ppp <- function(combined_ppp, main_title = "Combined PPP Plot") {
  # Plot the ppp object with title
  plot(combined_ppp, main = main_title)
}

# Helper function to compute the spatial KDE for ppp objects
compute_kde <- function(ppp_object, sigma_value = 1000) {
  # Compute the KDE using a Gaussian kernel
  kde_result <- density.ppp(ppp_object, sigma = sigma_value, edge = TRUE)
  return(kde_result)
}
```

### Loop Through Each Event Type List and Compute KDE

Next, We will loop through each event type list (`b_ppp_list`, `erv_ppp_list`, `vac_ppp_list`, `sd_ppp_list`) and compute the KDE layers for each `ppp` object.

```{r}
sigma_value <- 10

# Battles
b_kde_list <- list()
for (ppp_name in names(b_ppp_list)) {
  kde_result <- compute_kde(b_ppp_list[[ppp_name]], sigma_value)
  b_kde_list[[ppp_name]] <- kde_result
}

# Explosions/Remote Violence
erv_kde_list <- list()
for (ppp_name in names(erv_ppp_list)) {
  kde_result <- compute_kde(erv_ppp_list[[ppp_name]], sigma_value)
  erv_kde_list[[ppp_name]] <- kde_result
}

# Violence Against Civilians
vac_kde_list <- list()
for (ppp_name in names(vac_ppp_list)) {
  kde_result <- compute_kde(vac_ppp_list[[ppp_name]], sigma_value)
  vac_kde_list[[ppp_name]] <- kde_result
}

# Strategic Developments
sd_kde_list <- list()
for (ppp_name in names(sd_ppp_list)) {
  kde_result <- compute_kde(sd_ppp_list[[ppp_name]], sigma_value)
  sd_kde_list[[ppp_name]] <- kde_result
}
```

```{r}
for (ppp_name in names(b_ppp_list)) {
  combined_b_ppp <- combine_ppp_with_owin(ppp_name, b_ppp_list[[ppp_name]])
  plot_combined_ppp(combined_b_ppp, main_title = paste("Battles:", ppp_name))
}

# Explosions/Remote Violence
for (ppp_name in names(erv_ppp_list)) {
  combined_erv_ppp <- combine_ppp_with_owin(ppp_name, erv_ppp_list[[ppp_name]])
  plot_combined_ppp(combined_erv_ppp, main_title = paste("Explosions/Remote Violence:", ppp_name))
}

# Violence Against Civilians
for (ppp_name in names(vac_ppp_list)) {
  combined_vac_ppp <- combine_ppp_with_owin(ppp_name, vac_ppp_list[[ppp_name]])
  plot_combined_ppp(combined_vac_ppp, main_title = paste("Violence Against Civilians:", ppp_name))
}

# Strategic Developments
for (ppp_name in names(sd_ppp_list)) {
  combined_sd_ppp <- combine_ppp_with_owin(ppp_name, sd_ppp_list[[ppp_name]])
  plot_combined_ppp(combined_sd_ppp, main_title = paste("Strategic Developments:", ppp_name))
}
```

#### Spatio-Temporal KDE Helper Function

helper function to compute spatio-temporal KDE using the `sparr::spattemp.density` function, where we can specify the spatial (`h`) and temporal (`lambda`) bandwidths.

```{r}
quarter_time_mapping <- list(
  "2021_Q1" = 1,
  "2021_Q2" = 2,
  "2021_Q3" = 3,
  "2021_Q4" = 4,
  "2022_Q1" = 5,
  "2022_Q2" = 6,
  "2022_Q3" = 7,
  "2022_Q4" = 8,
  "2023_Q1" = 9,
  "2023_Q2" = 10,
  "2023_Q3" = 11,
  "2023_Q4" = 12,
  "2024_Q1" = 13,
  "2024_Q2" = 14
)


assign_temporal_marks <- function(ppp_object, quarter_string) {
  if (quarter_string %in% names(quarter_time_mapping)) {
    base_time_mark <- quarter_time_mapping[[quarter_string]]
    num_points <- npoints(ppp_object)
    time_variation <- runif(num_points, min = 0, max = 0.25)  # 

    ppp_object$marks <- base_time_mark + time_variation
  } else {
    stop("Quarter string not found in mapping")
  }
  return(ppp_object)
}



compute_spatio_temporal_kde <- function(ppp_object, h_value = 1000, lambda_value = 1) {
  # Ensure that the temporal marks exist and are numeric
  if (is.numeric(ppp_object$marks)) {
    time_range <- range(ppp_object$marks)
    print(time_range[1])
    print(time_range[2])
    
    # Ensure tlim is valid (tlim[1] < tlim[2])
    if (time_range[1] < time_range[2]) {
      kde_result <- spattemp.density(ppp_object, h = h_value, lambda = lambda_value, tlim = time_range)
      return(kde_result)
    } else {
      stop("Invalid temporal range: tlim[1] must be < tlim[2]")
    }
  } else {
    stop("Temporal marks are not numeric in the ppp object.")
  }
}
```

```{r}
combine_ppp_quarters <- function(ppp_list, quarter_labels) {
  all_points <- NULL
  for (i in seq_along(ppp_list)) {
    ppp_object <- ppp_list[[i]]
    
    x_coords <- ppp_object$x
    y_coords <- ppp_object$y
    n_points <- npoints(ppp_object)
    
    time_marks <- rep(quarter_labels[i], n_points)
    
    points_matrix <- cbind(x_coords, y_coords, time_marks)
    

    all_points <- rbind(all_points, points_matrix)
  }
  

  combined_ppp <- ppp(all_points[, 1], all_points[, 2], marks = all_points[, 3], 
                      window = ppp_list[[1]]$window, unitname = ppp_list[[1]]$unitname)
  
  return(combined_ppp)
}


quarter_labels <- c(1, 2, 3, 4) 
```

```{r}
b_ppp_list <- list(
  "myanmar_ppp_temp_km_2021_Q1_b" = get("myanmar_ppp_temp_km_2021_Q1_b"),
  "myanmar_ppp_temp_km_2021_Q2_b" = get("myanmar_ppp_temp_km_2021_Q2_b"),
  "myanmar_ppp_temp_km_2021_Q3_b" = get("myanmar_ppp_temp_km_2021_Q3_b"),
  "myanmar_ppp_temp_km_2021_Q4_b" = get("myanmar_ppp_temp_km_2021_Q4_b")
)
erv_ppp_list <- list(
  "myanmar_ppp_temp_km_2021_Q1_erv" = get("myanmar_ppp_temp_km_2021_Q1_erv"),
  "myanmar_ppp_temp_km_2021_Q2_erv" = get("myanmar_ppp_temp_km_2021_Q2_erv"),
  "myanmar_ppp_temp_km_2021_Q3_erv" = get("myanmar_ppp_temp_km_2021_Q3_erv"),
  "myanmar_ppp_temp_km_2021_Q4_erv" = get("myanmar_ppp_temp_km_2021_Q4_erv")
)

sd_ppp_list <- list(
  "myanmar_ppp_temp_km_2021_Q1_sd" = get("myanmar_ppp_temp_km_2021_Q1_sd"),
  "myanmar_ppp_temp_km_2021_Q2_sd" = get("myanmar_ppp_temp_km_2021_Q2_sd"),
  "myanmar_ppp_temp_km_2021_Q3_sd" = get("myanmar_ppp_temp_km_2021_Q3_sd"),
  "myanmar_ppp_temp_km_2021_Q4_sd" = get("myanmar_ppp_temp_km_2021_Q4_sd")
)

vac_ppp_list <- list(
  "myanmar_ppp_temp_km_2021_Q1_vac" = get("myanmar_ppp_temp_km_2021_Q1_vac"),
  "myanmar_ppp_temp_km_2021_Q2_vac" = get("myanmar_ppp_temp_km_2021_Q2_vac"),
  "myanmar_ppp_temp_km_2021_Q3_vac" = get("myanmar_ppp_temp_km_2021_Q3_vac"),
  "myanmar_ppp_temp_km_2021_Q4_vac" = get("myanmar_ppp_temp_km_2021_Q4_vac")
)


# Combine them and assign temporal marks
combined_b_ppp <- combine_ppp_quarters(b_ppp_list, quarter_labels)
kde_result_b <- spattemp.density(combined_b_ppp, h = 1000, lambda = 2)
plot(kde_result_b)
# Combine and compute KDE for Explosions/Remote Violence
combined_erv_ppp <- combine_ppp_quarters(erv_ppp_list, quarter_labels)
kde_result_erv <- spattemp.density(combined_erv_ppp, h = 1000, lambda = 2)
plot(kde_result_erv)

# Combine and compute KDE for Strategic Developments
combined_sd_ppp <- combine_ppp_quarters(sd_ppp_list, quarter_labels)
kde_result_sd <- spattemp.density(combined_sd_ppp, h = 1000, lambda = 2)
plot(kde_result_sd)

# Combine and compute KDE for Violence Against Civilians
combined_vac_ppp <- combine_ppp_quarters(vac_ppp_list, quarter_labels)
kde_result_vac <- spattemp.density(combined_vac_ppp, h = 1000, lambda = 2)
plot(kde_result_vac)
```

```{r}

helper_func_plot_kde_list <- function(kde_list, xlab = "Distance (km)", ylab = "Density") {
  for (name in names(kde_list)) {
    plot(kde_list[[name]], main = name, xlab = xlab, ylab = ylab)
  }
} 
  
kde_list_battles <- list()
kde_list_erv <- list()
kde_list_sd <- list()
kde_list_vac <- list()

# Combine and compute KDE for Battles
combined_b_ppp <- combine_ppp_quarters(b_ppp_list, quarter_labels)
kde_list_battles$Battles <- spattemp.density(combined_b_ppp, h = 1000, lambda = 2)

# Combine and compute KDE for Explosions/Remote Violence
combined_erv_ppp <- combine_ppp_quarters(erv_ppp_list, quarter_labels)
kde_list_erv$ExplosionsRemoteViolence <- spattemp.density(combined_erv_ppp, h = 1000, lambda = 2)

# Combine and compute KDE for Strategic Developments
combined_sd_ppp <- combine_ppp_quarters(sd_ppp_list, quarter_labels)
kde_list_sd$StrategicDevelopments <- spattemp.density(combined_sd_ppp, h = 1000, lambda = 2)

# Combine and compute KDE for Violence Against Civilians
combined_vac_ppp <- combine_ppp_quarters(vac_ppp_list, quarter_labels)
kde_list_vac$ViolenceAgainstCivilians <- spattemp.density(combined_vac_ppp, h = 1000, lambda = 2)

# Now use the helper_func_plot_kde_list for each event type

# Plot Battles KDE
helper_func_plot_kde_list(kde_list_battles, xlab = "Distance (km)", ylab = "Density")

# Plot Explosions/Remote Violence KDE
helper_func_plot_kde_list(kde_list_erv, xlab = "Distance (km)", ylab = "Density")

# Plot Strategic Developments KDE
helper_func_plot_kde_list(kde_list_sd, xlab = "Distance (km)", ylab = "Density")

# Plot Violence Against Civilians KDE
helper_func_plot_kde_list(kde_list_vac, xlab = "Distance (km)", ylab = "Density")
```

## **2nd-order spatial point pattern analysis**

On **spatial-temporal KDE layers**, typically i would use computing functions like the **Ripley’s K-function** or **L-function**. These functions help identify clustering patterns over different spatial scales.

Given that I'm working with **spatio-temporal KDE layers**, to conduct **2nd-order spatial point pattern analysis,** i will use `spatstat` functions on my earlier KDE results.

### Steps:

1.  **Convert KDE layers to spatial point patterns**.

2.  **Compute Ripley’s K-function (Kest)** or L-function (Lest).

3.  **Plot the results to observe clustering or dispersion**.

### Compute using K-function and/or L-function

```{r}

filter_ppp_by_year <- function(ppp_obj, year) {
  # Extract the year part from marks or events
  event_years <- marks(ppp_obj)
  
  # Filter the ppp object based on the provided year
  filtered_ppp <- ppp_obj[event_years == year]
  
  return(filtered_ppp)
}

# Helper function for K-function calculation and plotting
compute_k_function <- function(ppp_obj, years) {
  k_results <- list()
  
  for (year in years) {
    filtered_ppp <- filter_ppp_by_year(ppp_obj, year)
    
    if (npoints(filtered_ppp) > 0) {  # Ensure there's data for that year
      K_result <- Kest(filtered_ppp, correction = "Ripley")
      plot(K_result, . - r ~ r, main = paste("K-function for", year), xlim = c(0, 1000))
      k_results[[as.character(year)]] <- K_result
    }
  }
  
  return(k_results)
}
# Helper function for L-function calculation and plotting
compute_l_function <- function(ppp_obj, years) {
  l_results <- list()
  
  for (year in years) {
    filtered_ppp <- filter_ppp_by_year(ppp_obj, year)
    
    if (npoints(filtered_ppp) > 0) {  # Ensure there's data for that year
      L_result <- Lest(filtered_ppp, correction = "Ripley")
      plot(L_result, . - r ~ r, main = paste("L-function for", year), xlim = c(0, 1000))
      l_results[[as.character(year)]] <- L_result
    }
  }
  
  return(l_results)
}
```

```{r}
years_to_analyze <- c(2021, 2022, 2023, 2024)

# Battles (combined_b_ppp)
k_battles_results <- compute_k_function(combined_b_ppp, years_to_analyze)
l_battles_results <- compute_l_function(combined_b_ppp, years_to_analyze)

# Explosions/Remote Violence (combined_erv_ppp)
k_erv_results <- compute_k_function(combined_erv_ppp, years_to_analyze)
l_erv_results <- compute_l_function(combined_erv_ppp, years_to_analyze)

# Strategic Developments (combined_sd_ppp)
k_sd_results <- compute_k_function(combined_sd_ppp, years_to_analyze)
l_sd_results <- compute_l_function(combined_sd_ppp, years_to_analyze)

# Violence Against Civilians (combined_vac_ppp)
k_vac_results <- compute_k_function(combined_vac_ppp, years_to_analyze)
l_vac_results <- compute_l_function(combined_vac_ppp, years_to_analyze)
```

### Visualizing the Results

```{r}
# Plot K-function results for battles by year
helper_func_plot_kde_list(k_battles_results, xlab = "Distance (km)", ylab = "K(d)-r")

# Plot L-function results for strategic developments by year
helper_func_plot_kde_list(l_sd_results, xlab = "Distance (km)", ylab = "L(d)-r")
```
